Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-02-09 pkg/disttree/inst/draft/draft.Rnw
trigger rebuild
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2003 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e44297efc7abc249892430749c53aa42251febb0

In regression analysis various models have been developed to describe the
relationship between a response variable and a set of covariates. The classical
model is the linear model (LM) where the conditional mean of the response
is modeled through a linear function of the covariates (see the left panel of
Figure~\ref{Devel_parmod} for a schematic illustration). Over the last decades
this has been extended in various directions including:
\begin{itemize}
  \item Generalized linear models (GLMs, \citealp{McCullagh+Nelder:1989})
    encompassing an additional nonlinear link function for the conditional mean.
  \item Generalized additive models (GAMs, \citealp{Hastie+Tibshirani:1986})
    allowing for smooth nonlinear effects in the covariates
    (Figure~\ref{Devel_parmod}, middle).
  \item Generalized additive models for location, scale, and shape
    (GAMLSS, \citealt{stasinopoulos2005}) adopting a probabilistic modeling
    approach. In GAMLSS, each parameter of a statistical distribution can depend
    on an additive predictor of the covariates comprising linear and/or
    smooth nonlinear terms (Figure~\ref{Devel_parmod}, right).
\end{itemize}
Thus, the above-mentioned models provide a broad toolbox for capturing different
aspects of the response (mean only vs.\ full distribution) and different types
of dependencies on the covariates (linear vs.\ nonlinear additive terms).

@@

While in many applications conditional mean regression models had been receiving
the most attention, there has been a paradigm shift over the last decade towards
distributional regression models. An important reason for this is that in many
fields forecasts of the mean are not the only (or not even the main) concern but
that there is an increasing interest in probabilistic forecasts. Quantities of
interest typically include exceedence probabilities for certain thresholds of the
response or quantiles of the response distribution. As an example, consider
weather forecasting where there is less interest in the mean amount of
precipitation on the next day. Instead, the probability of rain vs.\ no rain
is typically more relevant or, in some situations, a prediction interval of
expected precipitation (say from the expected 10\% to 90\% quantile). Similar
considerations apply for other meteorological quantities and hence attention
in the weather forecasting literature has been shifting from classical linear
models \citep{Glahn+Lowry:1972} towards probabilistic models such as the
non-homogeneous Gaussian regression (NGR) of \cite{gneiting2005calibrated}.
The NGR typically describes the mean of some meteorological response variable through
the average of the corresponding quantity from an ensemble of physically-based
numerical weather predictions (NWPs). Similarly, the variance of the response
is captured through the variance of the ensemble of NWP outputs. Thus, the NGR
considers both the mean as well as the uncertainty of the NWP model outputs
to obtain probabilistic forecasts calibrated to a particular site.

@@

The so far discussed parametric models with the response depending on additive linear or 
smooth nonlinear terms of the covariates are powerful tools as long as only smooth effects 
are to be considered. A different approach that can also deal with non-additive effects and 
possibly unknown interactions of covariates is to apply tree-based methods. 
A regression tree (\citealt{breiman1984classification}) works on the idea of separating 
the data into more homogeneous subgroups based on covariates. However, the resulting step 
function, as illustrated in the left panel of Figure \ref{Devel_treeforest}, can often be too 
rough and impose abrupt changes on the model even in case of smooth transitions. Random 
forest models such as Breiman and Cutler's randomForest (\citealt{breiman2001random}) 
solve this problem by combining an ensemble of trees such that steps are smoothed 
out in case of smooth effects while abrupt changes are still captured by its tree-based 
structure (see the top middle panel of Figure \ref{Devel_treeforest}).

While classical regression trees and random forests only model the mean of the response
we are now going to combine the idea of these tree based-methods with the progress
made in the field of flexible parametric modeling. In particular, the development from mean regression
models to distributional models will be embedded in the framework of regression trees and
random forests providing distributional trees and forests. This development is illustrated
in the bottom panel of Figure~\ref{Devel_treeforest} for distributional trees and in the top right panel
for distributional forests.

>>>>

In regression analysis a wide range of models has been developed to describe the
relationship between a response variable and a set of covariates. The classical
model is the linear model (LM) where the conditional mean of the response
is modeled through a linear function of the covariates (see the left panel of
Figure~\ref{Devel_parmod} for a schematic illustration). Over the last decades
this has been extended in various directions including:
\begin{itemize}
  \item \emph{Generalized linear models} (GLMs, \citealp{Nelder+Wedderburn:1972})
    encompassing an additional nonlinear link function for the conditional mean.
  \item \emph{Generalized additive models} (GAMs, \citealp{Hastie+Tibshirani:1986})
    allowing for smooth nonlinear effects in the covariates
    (Figure~\ref{Devel_parmod}, middle).
  \item \emph{Generalized additive models for location, scale, and shape}
    (GAMLSS, \citealt{stasinopoulos2005}) adopting a probabilistic modeling
    approach. In GAMLSS, each parameter of a statistical distribution can depend
    on an additive predictor of the covariates comprising linear and/or
    smooth nonlinear terms (Figure~\ref{Devel_parmod}, right).
\end{itemize}
Thus, the above-mentioned models provide a broad toolbox for capturing different
aspects of the response (mean only vs.\ full distribution) and different types
of dependencies on the covariates (linear vs.\ nonlinear additive terms).

@@

While in many applications conditional mean regression models have been receiving
the most attention, there has been a paradigm shift over the last decade towards
distributional regression models. An important reason for this is that in many
fields forecasts of the mean are not the only (or not even the main) concern but
instead there is an increasing interest in probabilistic forecasts. Quantities of
interest typically include exceedence probabilities for certain thresholds of the
response or quantiles of the response distribution. As an example, consider
weather forecasting where there is less interest in the mean amount of
precipitation on the next day. Instead, the probability of rain vs.\ no rain
is typically more relevant or, in some situations, a prediction interval of
expected precipitation (say from the expected 10\% to 90\% quantile). Similar
considerations apply for other meteorological quantities and hence attention
in the weather forecasting literature has been shifting from classical linear
models \citep{Glahn+Lowry:1972} towards probabilistic models such as the
non-homogeneous Gaussian regression (NGR) of \cite{gneiting2005calibrated}.
The NGR typically describes the mean of some meteorological response variable through
the average of the corresponding quantity from an ensemble of physically-based
numerical weather predictions (NWPs). Similarly, the variance of the response
is captured through the variance of the ensemble of NWP outputs. Thus, the NGR
considers both the mean as well as the uncertainty of the NWP model outputs
to obtain probabilistic forecasts calibrated to a particular site.

@@

In summary, the models discussed so far provide a broad and powerful toolset for
parametric distributional fits depending on a specified set of additive linear or
smooth nonlinear terms. A rather different approach to capturing the dependence on
covariates are tree-based models.
\begin{itemize}
  \item \emph{Regression trees} (\citealt{breiman1984classification}) recursively
    split the data into more homogeneous subgroups and can thus capture abrupt shifts
    (Figure~\ref{Devel_treeforest}, top left) and approximate nonlinear functions.
    Furthermore, trees automatically carry out a forward selection of covariates and
    their interactions. 
  \item \emph{Random forests} (\citealt{breiman2001random}) average the predictions of
    an ensemble of trees fitted to resampled versions of the learning data. This
    stabilizes the recursive partitions from individual trees and hence better
    approximates smooth functions (Figure~\ref{Devel_treeforest}, top middle)
\end{itemize}
While classical regression trees and random forests only model the mean of the response
we propose to follow the ideas from (semi-)parametric regression modeling -- as outlined
in Figure~\ref{Devel_parmod} -- and combine tree-based methods with parametric distributional models,
yielding two novel techniques:
\begin{itemize}
  \item \emph{Distributional trees} split the data into more homogenous groups with
    respect to a parametric distribution, thus capturing changes in any distribution
    paramer like location, scale, or shape (Figure~\ref{Devel_treeforest}, bottom left)
  \item \emph{Distributional forests} utilize an ensemble of distributional trees
    for obtaining stabilized and smoothed parametric predictions (Figure~\ref{Devel_treeforest},
    top right).
\end{itemize}
